---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# spatialising

<!-- badges: start -->
[![Codecov test coverage](https://codecov.io/gh/Nowosad/spatialising/branch/main/graph/badge.svg)](https://app.codecov.io/gh/Nowosad/spatialising?branch=main)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![R-CMD-check](https://github.com/Nowosad/spatialising/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/Nowosad/spatialising/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

The goal of spatialising is to perform simulations of spatial raster data using the Ising model.

## Installation

You can install the development version of spatialising from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("Nowosad/spatialising")
```

## Example

```{r}
library(terra)
library(spatialising)
r1 = rast(system.file("raster/r_start.tif", package = "spatialising"))
plot(r1)
```

```{r}
r2 = spatial_ising(r1, B = -0.3, J = 0.7)
plot(r2)
```

```{r}
ri1 = spatial_ising(r1, B = -0.3, J = 0.7, timesteps = 9)
plot(ri1)
```

```{r}
ri2 = spatial_ising(r1, B = 0.3, J = 0.7, timesteps = 9)
plot(ri2)
```

```{r}
ri3 = spatial_ising(r1, B = -0.3, J = 0.4, timesteps = 9)
plot(ri3)
```


## Contribution

Contributions to this package are welcome - let us know if you have any suggestions or spotted a bug. 
The preferred method of contribution is through a GitHub pull request. 
Feel also free to contact us by creating [an issue](https://github.com/nowosad/spatialising/issues).
